---
title: "Project 1: Multivariate Non-Normal Distributions and Correlated Data"
output: html_document
---

# 1. Plan and specify distributions and correlation

We simulate medical data with heavy-tailed a Multivariate t-Distribution using a t-Copula to generate correlated variables: Glucose, Cholesterol, BMI, and Blood Pressure).  


```{r load_packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(MASS)  
library(Matrix)
library(copula)
library(mvtnorm)
library(ggplot2) 
library(GGally)
library(DT)
library(patchwork)
```


```{r set_parameters, message=FALSE}
set.seed(2025)

# Define parameters
nu <- 3  # Degrees of freedom (controls tail thickness)
p <- 4   # Number of variables
n <- 1000  # Number of samples

# Define mean vector
mu <- c(100,   # Mean Glucose Level (mg/dL)
        200,   # Mean Cholesterol Level (mg/dL)
        25,    # Mean BMI (Body Mass Index)
        120)   # Mean Blood Pressure (mmHg)
```

- Degrees of Freedom (nu = 5)

Lower degrees of freedom results in a heavier tail, which better simulates extreme cases in medical data (diabetes, obesity, etc.).  


- Number of Variables (p = 4)

These four variables (Glucose, Cholesterol, BMI, Blood Pressure) are commonly used metabolic and cardiovascular health indicators and are correlated to each other.

- Choice of Mean Values (mu)  
The mean values are set based on typical values of a healthy adult.


```{r corr_structure}
# Define block structures
# Block-diagonal: high correlation within groups, no correlation across groups

# Glucose & Cholesterol (Metabolism)
block1 <- matrix(c(1, 0.5, 
                   0.5, 1), nrow=2, ncol=2) 
# BMI & Blood Pressure (Cardiovascular)
block2 <- matrix(c(1, 0.4, 
                   0.4, 1), nrow=2, ncol=2)  

# Combine blocks into a block-diagonal matrix
Sigma <- bdiag(block1, block2)

# Convert to a standard matrix format
# Ensure the correlation matrix is positive definite
Sigma <- as.matrix(nearPD(Sigma)$mat) 
```

Copulas are the mechanism which allows us to isolate the dependency structure in a multivariate distribution. In particular, we can construct any multivariate distribution by separately specifying the marginal distributions and the copula. (Haugh, 2016) https://www.columbia.edu/~mh2078/QRM/Copulas.pdf

Idea behind copulas:

- It is a mathematical object that has input as 2 or more marginal distributions and has output as a joint  distribution

- If we have two probabilities that live on 0 -> 1, we can transform them into a higher space of 0 to infinity. Then when we add them together, they still live in the space of 0 to infinity. We take the inverse transformation of the sum, so the sum of the two probabilities will live on 0 -> 1 again.

- If we use a negative log transformation (which will put any probability from 0 to 1 on a scale from 0 to infinity), the inverse will be negative exponential transformation (which brings anything from 0 to infinity back to 0 to 1).

- If the two probabilities are independent, the joint probability is equal to the product of the individual probabilities (Frechet-Hoffding boundary copula). We're not so interested in it because we want to use copulas for distributions with correlation.

- Copula allows distributions that have dependency to have flexibility. e.g. Gumbel copula - dependency increases with extreme positive values, which captures the dependency better than any multivariate distribution.

- Gaussian copula transforms variables to standard normal distributions (mean = 0, variance = 1) by mapping them on a percentile-to-percentile basis. Those distributions are then combined to form a joint distribution 

- The t-copula is derived from the multivariate t-distribution, just as the Gaussian copula comes from the multivariate normal. t-copula better captures the phenomenon of dependent extreme values (tail dependence)

- Tail dependence is when the correlation between two variables increases as you get "further" in the tail (either or both) of the distribution. Gaussian copula has no tail dependence.

----------------------------to delete------------------------------
Medical Justification for Grouping Variables:  

1. Glucose & Cholesterol (Metabolic Block)  
These two variables are often correlated because they are both regulated by insulin and metabolic processes.  
Patients with high glucose levels (e.g., in diabetes) often have elevated cholesterol due to metabolic dysfunction.  

2. BMI & Blood Pressure (Cardiovascular Block)  
Higher BMI is strongly associated with increased blood pressure, as obesity contributes to hypertension and cardiovascular disease.  
This correlation reflects real-world epidemiological findings in obesity and hypertension studies.  

```{r t_copula_generation}
# Extract lower triangle correlations
rho_vec <- Sigma[lower.tri(Sigma)]  

# Define the t-Copula
copula_t <- tCopula(param=rho_vec, dim=p, df=nu, dispstr="un")

# Generate Copula-Based Uniform Samples
U_t <- rCopula(1000, copula_t) 

# Define upper and lower bounds for each variable
glucose_min <- 32  # minimum plausible fasting glucose level in mg/dL
glucose_max <- 500 # maximum plausible fasting glucose level in mg/dL

cholesterol_min <- 100  # minimum plausible cholesterol level in mg/dL
cholesterol_max <- 300  # maximum plausible cholesterol level in mg/dL

BMI_min <- 15    # minimum plausible BMI
BMI_max <- 40    # maximum plausible BMI

blood_pressure_min <- 90   # minimum plausible systolic blood pressure in mmHg
blood_pressure_max <- 180  # maximum plausible systolic blood pressure in mmHg

# Generate Independent Heavy-Tailed t-Distributions
glucose_raw <- scale(rt(10000, df=nu)) * 15 + mu[1]
cholesterol_raw <- scale(rt(10000, df=nu)) * 25 + mu[2]
BMI_raw <- scale(rt(10000, df=nu)) * 3 + mu[3]
blood_pressure_raw <- scale(rt(10000, df=nu)) * 10 + mu[4]

# Filter out values that fall outside the realistic human limits
glucose <- glucose_raw[glucose_raw >= glucose_min & glucose_raw <= glucose_max]
cholesterol <- cholesterol_raw[cholesterol_raw >= cholesterol_min & cholesterol_raw <= cholesterol_max]
BMI <- BMI_raw[BMI_raw >= BMI_min & BMI_raw <= BMI_max]
blood_pressure <- blood_pressure_raw[blood_pressure_raw >= blood_pressure_min & blood_pressure_raw <= blood_pressure_max]


# Map Copula Uniform Samples Back to Heavy-Tailed t-Distributions
data_copula_t <- data.frame(
  Glucose = quantile(glucose, probs = U_t[,1]),
  Cholesterol = quantile(cholesterol, probs = U_t[,2]),
  BMI = quantile(BMI, probs = U_t[,3]),
  BloodPressure = quantile(blood_pressure, probs = U_t[,4])
)

# Generate correlation matrix
corr_matrix <- cor(data_copula_t) %>% as.data.frame()
knitr::kable(corr_matrix, caption = "Correlation Matrix of t_copula", digits = 3)
```

----------------------------to delete------------------------------

A t-Copula is ideal for modeling correlated medical data with heavy tails, capturing realistic dependencies between variables.  

Medical variables (glucose, cholesterol, BMI, blood pressure) often show heavy-tailed distributions.  
The t-Copula ensures that extreme values tend to co-occur, reflecting real-world patterns (e.g., high glucose & high cholesterol in diabetes).  

A Gaussian Copula assumes weak tail dependence, failing to model joint extremes properly.  

```{r visualize_t_copula}
ggpairs(data_copula_t, 
        title = "t-Copula: Multivariate t-Distribution",
        lower = list(continuous = "points"),
        diag = list(continuous = "densityDiag"),
        upper = list(continuous = "cor")) +
  theme(plot.title = element_text(hjust = 0.5, face="bold", size=14))
```



# 2. Implement your data generation method

```{r generate_data}
# Function for generate multivariate t-distribution data 
generate_t_copula_data <- function(n = n, nu = nu) {
  
  # Define block-diagonal correlation structure
  block1 <- matrix(c(1, 0.5, 
                     0.5, 1), nrow=2, ncol=2)  # Metabolic Block (Glucose & Cholesterol)
  block2 <- matrix(c(1, 0.4, 
                     0.4, 1), nrow=2, ncol=2)  # Cardiovascular Block (BMI & Blood Pressure)

  # Combine blocks into a correlation matrix
  Sigma <- bdiag(block1, block2)
  Sigma <- as.matrix(nearPD(Sigma)$mat)  # Ensure it is positive definite

  # Extract lower triangle correlations for t-Copula
  rho_vec <- Sigma[lower.tri(Sigma)]  

  # Define the t-Copula
  copula_t <- tCopula(param=rho_vec, dim=p, df=nu, dispstr="un")

  # Generate Copula-Based Uniform Samples
  U_t <- rCopula(n, copula_t) 

  # Define upper and lower bounds for each variable
  glucose_min <- 32  # minimum plausible fasting glucose level in mg/dL
  glucose_max <- 500 # maximum plausible fasting glucose level in mg/dL
  
  cholesterol_min <- 100  # minimum plausible cholesterol level in mg/dL
  cholesterol_max <- 300  # maximum plausible cholesterol level in mg/dL
  
  BMI_min <- 15    # minimum plausible BMI
  BMI_max <- 40    # maximum plausible BMI
  
  blood_pressure_min <- 90   # minimum plausible systolic blood pressure in mmHg
  blood_pressure_max <- 180  # maximum plausible systolic blood pressure in mmHg
  
  # Generate Independent Heavy-Tailed t-Distributions
  glucose_raw <- scale(rt(10000, df=nu)) * 15 + mu[1]
  cholesterol_raw <- scale(rt(10000, df=nu)) * 25 + mu[2]
  BMI_raw <- scale(rt(10000, df=nu)) * 3 + mu[3]
  blood_pressure_raw <- scale(rt(10000, df=nu)) * 10 + mu[4]
  
  # Filter out values that fall outside the realistic human limits
  glucose <- glucose_raw[glucose_raw >= glucose_min & 
                           glucose_raw <= glucose_max]
  cholesterol <- cholesterol_raw[cholesterol_raw >= cholesterol_min 
                                 & cholesterol_raw <= cholesterol_max]
  BMI <- BMI_raw[BMI_raw >= BMI_min & BMI_raw <= BMI_max]
  blood_pressure <- blood_pressure_raw[blood_pressure_raw 
                                       >= blood_pressure_min & 
                                         blood_pressure_raw <= 
                                         blood_pressure_max]
  
  
  # Map Copula Uniform Samples Back to Heavy-Tailed t-Distributions
  data_copula_t <- data.frame(
    Glucose = quantile(glucose, probs = U_t[,1]),
    Cholesterol = quantile(cholesterol, probs = U_t[,2]),
    BMI = quantile(BMI, probs = U_t[,3]),
    BloodPressure = quantile(blood_pressure, probs = U_t[,4])
  )

  return(data_copula_t)
}

# Generate a dataset with 1000 samples and df = 5
generated_data <- generate_t_copula_data(n = n, nu = nu)

# Save data as CSV
write.csv(generated_data, "generated_t_copula_data.csv", row.names = FALSE)
```


```{r visualization_hist_fit}
plot_distribution <- function(data, variable_name, df=5) {
  x_min <- min(data[[variable_name]])
  x_max <- max(data[[variable_name]])
  mean_val <- mean(data[[variable_name]])
  sd_val <- sd(data[[variable_name]])
  
  ggplot(data, aes(x = .data[[variable_name]])) +  
    geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.7) +  
    stat_function(fun = function(x) dt((x - mean_val) / sd_val, df = df) / sd_val, 
                  col = "red", linewidth = .8) +  
    labs(title = paste("Histogram & t-Dist Fit for", variable_name), 
         x = variable_name, y = "Density") +
    theme_minimal() + 
    theme(plot.title = element_text(hjust = 0))
}

p1 <- plot_distribution(generated_data, "Glucose")
p2 <- plot_distribution(generated_data, "Cholesterol")
p3 <- plot_distribution(generated_data, "BMI")
p4 <- plot_distribution(generated_data, "BloodPressure")

(p1 + p2) / (p3 + p4)
```

This visualization compares the empirical distributions of the generated data (blue histograms) with the fitted t-distributions (red curves).  

The t-distribution fits well with the generated data with heavy-tailed distributions both at the center and on the tails.

Overall, the t-Copula effectively models correlations and heavy tails, making it a suitable choice for realistic medical data simulation.  


```{r}
# Compute the empirical correlation matrix
empirical_corr <- cor(data_copula_t)

# Display side-by-side comparison
comparison_matrix <- data.frame(
  Variable_1 = c("Glucose", "Glucose", "Glucose", 
                 "Cholesterol", "Cholesterol", "BMI"),
  Variable_2 = c("Cholesterol", "BMI", "Blood Pressure", 
                 "BMI", "Blood Pressure", "Blood Pressure"),
  Target_Correlation = c(0.5, 0, 0, 0, 0, 0.4),  # From defined Sigma
  Empirical_Correlation = c(
    empirical_corr["Glucose", "Cholesterol"],
    empirical_corr["Glucose", "BMI"],
    empirical_corr["Glucose", "BloodPressure"],
    empirical_corr["Cholesterol", "BMI"],
    empirical_corr["Cholesterol", "BloodPressure"],
    empirical_corr["BMI", "BloodPressure"]
  )
)

# Display the table
knitr::kable(comparison_matrix, 
             caption = "Comparison of Target vs. Empirical Correlations", 
             digits = 3)
```

