---
title: "simstudy_2"
author: "Mirah"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Design Simulation study 2

Identify a statistical inference tool that could accommodate the correlation and non-normality ((e.g., a robust regression approach, generalized estimating equations (GEE) for correlated data, or a Bayesian method with appropriate priors).

Use simulation that is similar to Simulation 1 to this inference tool provides valid results for non-normal, correlated data, and compare these results to the standard inference tool in study 1.

For this simulation, robust regression was used to mitigate the influence of outliers and heavy-tailed distributions. It is useful when ordinary least squares assumptions don't hold up Specifically, MM-estimation was performed because this method is useful when handeling correlated and non-normal data. The model works by using two steps. The first step is an initial robust estimation with $\beta$ of high resistance to outliers, and the second step involves refining the estimate using a robust lost function. This form of regression downweigns the influence of outliers and works well with skewed data. MM-estimation robust regressing has a high breakdown point compared to other regression models, meaning a high fraction of outliers for which a model won't break down.
```{r}
library(tidyverse)
library(ggplot2)
library(robustbase)
library(copula)
library(Matrix)
```

```{r}
#taken from parts 1 and 2 and turned into a function so I can change parapeters in the simulation

#function to generate correlated non-normal data using a t-Copula (used part 2)
generate_t_copula_data <- function(n, nu, rho_mod = NULL) {
  #block-diagonal correlation structure
  block1 <- matrix(c(1, 0.5, 0.5, 1), 2, 2)  #metabolic block
  block2 <- matrix(c(1, 0.4, 0.4, 1), 2, 2)  #cardiovascular block
  
  #modify correlation for later
  if (!is.null(rho_mod)) {
    block1[1,2] <- block1[2,1] <- rho_mod[1]
    block2[1,2] <- block2[2,1] <- rho_mod[2]
  }
  
  #correlation matrix
  Sigma <- bdiag(block1, block2)
  Sigma <- as.matrix(nearPD(Sigma)$mat)
  rho_vec <- Sigma[lower.tri(Sigma)]  
  
  #define the t-Copula
  copula_t <- tCopula(param=rho_vec, dim=4, df=nu, dispstr="un")
  U_t <- rCopula(n, copula_t)  
  
  #generate independent heavy-tailed distributions
  mu <- c(100, 200, 25, 120) #the means
  
# Define upper and lower bounds for each variable
glucose_min <- 32  # minimum plausible fasting glucose level in mg/dL
glucose_max <- 500 # maximum plausible fasting glucose level in mg/dL

cholesterol_min <- 100  # minimum plausible cholesterol level in mg/dL
cholesterol_max <- 300  # maximum plausible cholesterol level in mg/dL

BMI_min <- 15    # minimum plausible BMI
BMI_max <- 40    # maximum plausible BMI

blood_pressure_min <- 90   # minimum plausible systolic blood pressure in mmHg
blood_pressure_max <- 180  # maximum plausible systolic blood pressure in mmHg

  glucose_raw <- scale(rt(10000, df=nu)) * 15 + mu[1]
  cholesterol_raw <- scale(rt(10000, df=nu)) * 25 + mu[2]
  bmi_raw <- scale(rt(10000, df=nu)) * 3 + mu[3]
  blood_pressure_raw <- scale(rt(10000, df=nu)) * 10 + mu[4]
  
  # Filter out values that fall outside the realistic human limits
glucose <- glucose_raw[glucose_raw >= glucose_min & glucose_raw <= glucose_max]
cholesterol <- cholesterol_raw[cholesterol_raw >= cholesterol_min & cholesterol_raw <= cholesterol_max]
bmi <- bmi_raw[bmi_raw >= BMI_min & bmi_raw <= BMI_max]
blood_pressure <- blood_pressure_raw[blood_pressure_raw >= blood_pressure_min & blood_pressure_raw <= blood_pressure_max]

  #map Copula Uniform Samples Back to Heavy-Tailed Distributions
  data_copula_t <- data.frame(
    glucose = quantile(glucose, probs = U_t[,1]),
    cholesterol = quantile(cholesterol, probs = U_t[,2]),
    bmi = quantile(bmi, probs = U_t[,3]),
    blood_pressure = quantile(blood_pressure, probs = U_t[,4])
  )
  return(data_copula_t)
}

```
$$
\hat{\beta} = \arg\min_{\beta} \sum_{i=1}^{n} \rho \left( \frac{y_i - x_i^\top \beta}{\hat{\sigma}} \right)
$$
$y_i$ : dependent variable
$x_i$ : predictor variable
$\beta$ : regression coefficients
$\sigma$ : robust estimate of residual standard deviation
$\rho$: Robust loss function

Original Data & Robust Regression
```{r}
set.seed(2025)

original_data <- generate_t_copula_data(n = 100, nu = 5)
original_data <- original_data %>%
  mutate(across(everything(), ~ . - mean(.)))#centering the data. this prevents the intercept variance from over affecting the data 
original_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = original_data)
summary(original_model) #robust regression using MM
```
Larger Sample Size

```{r}
set.seed(2025)

large_data <- generate_t_copula_data(n = 500, nu = 5)
large_data <- large_data %>%
  mutate(across(everything(), ~ . - mean(.)))
large_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = large_data)
summary(large_model)
```
Varying Correlation Levels

```{r}
set.seed(2025)
high_corr_data <- generate_t_copula_data(n = 100, nu = 5, rho_mod = c(0.8, 0.7))
high_corr_data <- high_corr_data %>%
  mutate(across(everything(), ~ . - mean(.)))
high_corr_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = high_corr_data)
summary(high_corr_model)

low_corr_data <- generate_t_copula_data(n = 100, nu = 5, rho_mod = c(0.2, 0.1))
low_corr_data <- low_corr_data %>%
  mutate(across(everything(), ~ . - mean(.)))
low_corr_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = low_corr_data)
summary(low_corr_model)
```

Varying Non-Normality Degree
```{r}
low_df_data <- generate_t_copula_data(n = 100, nu = 4)
low_df_data <-low_df_data %>%
  mutate(across(everything(), ~ . - mean(.)))
low_df_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = low_df_data)
summary(low_df_model)

high_df_data <- generate_t_copula_data(n = 100, nu = 6)
high_df_data <- high_df_data %>%
  mutate(across(everything(), ~ . - mean(.)))
high_df_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = high_df_data)
summary(high_df_model)

```
Lower than 4, and the data did not converge
```{r}
set.seed(2025)
bmi_data <- generate_t_copula_data(n = 100, nu = 5)

# Add a strong linear effect of bmi on glucose
bmi_data$glucose <- bmi_data$glucose + 2 * bmi_data$bmi  # Adjust the coefficient (0.5) as needed

#center
bmi_data <- bmi_data %>% mutate(across(everything(), ~ . - mean(.)))

#robust regression model
bmi_model <- lmrob(glucose ~ blood_pressure + cholesterol + bmi, data = bmi_data)
summary(bmi_model)
```

Visualizing results

```{r}

true_mean_glucose <- 0 #because data was centered around 0

#variance
calculate_variance <- function(model) {
  return(diag(vcov(model)))
}

#bias for glucose 
calculate_bias_glucose <- function(model, true_mean_glucose, data) {
  predicted_mean_glucose <- predict(model, newdata = data.frame(
    cholesterol = 0,  # all centered
    bmi = 0,          
    blood_pressure = 0
  ))
  
  #bias = predicted mean - true mean
  bias <- predicted_mean_glucose - true_mean_glucose
  return(bias)
}

#store results
store_results <- function(model, scenario_name, true_mean_glucose, data) {
  variance <- calculate_variance(model)
  bias <- calculate_bias_glucose(model, true_mean_glucose, data)
  
  data.frame(
    Scenario = scenario_name,
    Variance_Intercept = variance[1],
    Variance_BloodPressure = variance[2],
    Variance_Cholesterol = variance[3],
    Variance_BMI = variance[4],
    Bias_Glucose = bias
  )
}

#store results for each scenario
results <- rbind(
  store_results(original_model, "Original Data", true_mean_glucose, original_data),
  store_results(large_model, "Larger Sample Size", true_mean_glucose, large_data),
  store_results(high_corr_model, "High Correlation Levels", true_mean_glucose, high_corr_data),
  store_results(low_corr_model, "Low Correlation Levels", true_mean_glucose, low_corr_data),
  store_results(high_df_model, "High df (Less Non-Normal)", true_mean_glucose, high_df_data),
  store_results(low_df_model, "Low df (More Non-Normal)", true_mean_glucose, low_df_data),
  store_results(bmi_model, "BMI Effect Model", true_mean_glucose, bmi_data)
)

#results table
kable(results, caption = "Comparison of Variance and Bias Across Different Scenarios")
```


```{r}
#variance (for coefficients)
variance_long <- results %>%
  pivot_longer(cols = starts_with("Variance"), 
               names_to = "Coefficient", 
               values_to = "Variance")

ggplot(variance_long, aes(x = Scenario, y = Variance, fill = Coefficient)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Variance of Coefficients", x = "Scenario", y = "Variance") +
  theme_minimal()
```


```{r}
#bias (for glucose)
ggplot(results, aes(x = Scenario, y = Bias_Glucose, fill = Scenario)) +
  geom_bar(stat = "identity") +
  labs(title = "Bias of Glucose Predictions", x = "Scenario", y = "Bias") +
  theme_minimal()
```

```{r}
library(dplyr)
library(knitr)

# Function to extract R^2 and significant predictors
extract_results <- function(model, scenario_name) {
  # Extract R^2
  r_squared <- summary(model)$r.squared
  
  # Extract coefficients and p-values
  coef_table <- summary(model)$coefficients
  significant_predictors <- rownames(coef_table)[coef_table[, "Pr(>|t|)"] < 0.05]
  
  # Remove intercept from significant predictors
  significant_predictors <- significant_predictors[significant_predictors != "(Intercept)"]
  
  # Combine into a data frame
  data.frame(
    Scenario = scenario_name,
    R_Squared = r_squared,
    Significant_Predictors = paste(significant_predictors, collapse = ", ")
  )
}

results_table <- rbind(
  extract_results(original_model, "Original Data"),
  extract_results(large_model, "Larger Sample Size"),
  extract_results(high_corr_model, "High Correlation Levels"),
  extract_results(low_corr_model, "Low Correlation Levels"),
  extract_results(high_df_model, "High df (Less Non-Normal)"),
  extract_results(low_df_model, "Low df (More Non-Normal)"),
  extract_results(bmi_model, "BMI Effect")
)


kable(results_table, caption = "Model Performance and Significant Predictors")
```


```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```

## Refrences

http://dx.doi.org/10.12732/ijpam.v91i3.7
